Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-04-06 manuscript.tex
Updates to tb1 manuscript
Figures, tables, revamped discussion, info. on other loci involved in
tillering

\documentclass[12pt]{article}
\usepackage{geometry} 
\usepackage{hyperref}
\usepackage{color}
\usepackage[pdftex]{graphicx}  
%\usepackage{caption}
\usepackage{natbib}
\usepackage{authblk}
\renewcommand{\baselinestretch}{1.5}
\bibliographystyle{apalike}

\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} }

@@

\author[1]{Laura Vann}
\author[1]{Thomas Kono}
\author[1]{Tanja Pyh\"aj\"arvi}
\author[1]{Matthew Hufford}
\author[1,2]{Jeffrey Ross-Ibarra\thanks{rossibarra@ucdavis.edu}}
\affil[1]{Department of Plant Sciences, University of California Davis}
\affil[2]{Center for Population Biology and Genome Center, University of California Davis}

@@

\jri{Please add addresses for all authors.  All should have plant sciences, but Kono, Tanja, and Huff need their new addresses too. }

@@

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (Hopscotch) upstream of tb1 enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the Hopscotch element, led us to investigate the role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the Hopscotch element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{parviglumis} and \emph{mexicana}. Analysis of linkage disequilibrium between the Hopscotch element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection.  Finally, two greenhouse experiments fail to find an important role for \emph{tb1} in controlling tillering in natural populations of \emph{parvilgumis}. \jri{Needs a concluding sentence. "Thus, we think..." or something.}

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \cite{Ross-Ibarra et al 2007}. Perhaps the central focus of the study of domestication has been the identification of genetic variation underlying agronomically important traits such as fruit size and plant architecture \cite{OlsenGross2010}. \jri{I believe you need to remove spaces from the bibtex citation names.  See how this OlsenGross2010 works but the others don't? should be able to work with whitespace but i can't figure out why not.}  Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \cite{Kovach and McCouch 2008}. But while some effort has been invested into understanding how wild alleles behave in their domesticated relatives, very little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors.

% SFS figure
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=150mm]{SFS.png}
    \caption{This is an example figure.} 
\label{sfs_non_syn}
  \end{center}
\end{figure*}
%-------------------------------------------------------------------

\jri{This is how you cite the figure in the text. "I want to show you all Figure \ref{sfs_non_syn}." Latex takes care of the numbering.  Please add all the figures to the text, and add them in in the middle like this close to where they are referenced. Note that it puts the figure where it thinks best, which is not necessarily near this text.}

Maize (\emph{Zea mays} ssp. \emph{mays}) was domesticated from the teosinte \emph{Zea mays} ssp. \emph{parviglumis} (hereafter, \emph{parviglumis}) roughly 10,000 B.P. in southwest Mexico \cite{Piperno et al 2009, Matsuoka et al 2002}. Domesticated maize and the teosintes are an attractive system in which to study domestication due to the abundance of genetic tools developed for maize and well-characterized domestication loci \cite{Hufford et al 2012a, Doebley 2004, Hufford et al 2012b}. Additionally, large naturally occurring populations of both \emph{Zea mays} ssp. \emph{parviglumis} (the wild progenitor of maize) and \emph{Zea mays} ssp. \emph{mexicana} (highland teosinte) can be found throughout Mexico \cite{Wilkes 1977, Hufford et al 2013}, and genetic diversity and effective population size (Ne) of these taxa is estimated to be high \cite{Ross-Ibarra et al 2009} relative to other plants \cite{Nybom 2004}.

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis for these changes has been a focus of maize research for a number of years. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \cite{Doebley et al 1990, Doebley and Stec 1991, Lukens and Doebley 1999}. Further studies have shown that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer et al 2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \cite{Studer et al 2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \cite{Studer et al 2011}. Furthermore, \cite{Studer and Doebley 2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{mexicana}, \emph{parviglumis}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

@@

We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 Zea mays ssp. mexicana (hereafter mexicana) populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the Hopscotch insertion (Table S1, Table S2). \jri{please add tables and reference same as figs; see example in Sofiane's "deleterious 282" repo on lab github page} DNA was extracted from leaf tissue using a modified CTAB approach \cite{Doyle and Doyle 1990, Maloof et al 1984}. We designed primers using PRIMER3 \cite{Rozen and Skaletsky 2000} implemented in Geneious \cite{Kearse et al 2012} to amplify the entire Hopscotch element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence (~5 kb amplification product) and absence (~300 bp amplification product) of the Hopscotch insertion. Two PCRs were performed for each individual, one with primers flanking the Hopscotch (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, 5'-TCGTTGATGCTTTGATGGATGG-3'; 
Hop R, 5'-AACAGTATGATTTCATGGGACCG-3'; and HopIntR, 5'-CCTCCACCTCTCATGAGATCC-3' (Figure S1). \jri{supplemental material should be added to this file too.} Homozygotes show a single band for either the Hopscotch element (~5 kb) or the absence of the element (~300 bp), while heterozygotes are three-banded, showing a band for both the presence and absence of the Hopscotch element as well as a band for the internal primer set (Figure S2). When only one PCR resolved well, we scored one allele for the individual, which explains the odd number of alleles included in our analyses. We used Phusion High Fidelity Enzyme (Finnzymes, Inc.) and the following conditions for amplifications: 98°C for 3 min, 30 cycles of 98°C for 15 s, 65°C for 30 s, and 72°C for 3 min 30 s, with a final extension of 72°C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the Hopscotch based on band size.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \cite{Wang et al 1999, Clark et al 2006}. We designed the following primers using PRIMER3 \cite{Rozen and Skaletsky 2000}: for the 5' UTR, 5' GGATAATGTGCACCAGGTGT 3' and 5' GCGTGCTAGAGACACYTGTTGCT 3'; for the 50 kb upstream region, 5' TGTCCTCGCCGCAACTC 3' and 5' TGTACGCCCGCCCCTCATCA 3' (Figure S1). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94°C for 3 min, 30 cycles of 92°C for 40 s, annealing for 1 min, 72°C for 40 s, and a final 10 min extension at 72°C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7°C and 58.8°C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 µL of amplification product. This mix was placed on a thermal cycler with the following program: 37°C for 30 min, 80°C for 15 min, and a final cool-down step to 4°C.

We cloned cleaned fragments into a TOPO-TA vector (Invitrogen, Carlsbad) using OneShot TOP10 chemically competent E. coli cells, with an extended ligation time of 30 min for a complex target fragment. We plated cells on LB agar plates containing kanamycin, and screened colonies using vector primers M13 Forward and M13 Reverse under the following conditions: 96°C for 5 min; then 35 cycles at 96°C for 30 s, 53°C for 30 s, 72°C for two min; and a final extension at 72°C for 4 min. We visualized amplification products for incorporation of our insert on a 1\% agarose TAE gel.

Amplification products with successful incorporation of our insert were cleaned using Exonuclease 1 and Antarctic Phosphatase following the procedures detailed above, and sequenced with vector primers M13 Forward and M13 Reverse using Sanger sequencing at the College of Agriculture and Environmental Sciences (CAES) sequencing center at UC Davis. We aligned and trimmed primer sequences from resulting sequences using the software Geneious \cite{Kearse et al 2012}. Following alignment, we verified singleton SNPs by sequencing an additional one to four colonies from each clone. If the singleton was not present in these additional sequences it was considered an amplification or cloning error, and we replaced the base with the base of the additional sequences. If the singleton appeared in at least one of the additional sequences we considered it a real variant and kept it for further analyses.

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE), and to calculate differentiation between populations (FST) and subspecies (FCT) we used HierFstat \cite{Goudet 2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \cite{Coop et al 2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\cite{Coop et al 2010}. We used genotyping and covariance data from Pyhäjärvi et al., (2013) \cite{Pyhajarvi et al 2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (Table S3). Environmental data were obtained from www.worldclim.org, the Harmonized World Soil Database and www.harvestchoice.org, and summarized by principle component analysis Pyhäjärvi et al. (2013) \cite{Pyhajarvi et al 2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \cite{Thornton 2003} to calculate pairwise FST between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator$\theta_{W}$; pairwise nucleotide diversity $\theta_{\pi}$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (http://evolution.genetics.washington.edu/phylip.html) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \cite{Chia et al 2012}).

In order to assess patterns of linkage disequilibrium (LD) around the Hopscotch element in the context of chromosomal patterns of LD we used Tassel \cite{Bradbury et al 2007} and calculated LD between SNPs across chromosome 1 using previously published data from twelve plants each of the Ejutla A, Ejutla B, San Lorenzo, and La Mesa populations \cite{Pyhajarvi et al 2013}. We chose these populations because we had both genotyping data for the Hopscotch as well as chromosome-wide SNP data for chromosome 1. For each population we filtered the initial set of 5,897 SNPs on chromosome 1 to accept only SNPs with a minor allele frequency of at least 0.1, resulting in 1,671, 3,023, 3,122, and 2,167 SNPs for San Lorenzo, Ejutla B, Ejutla A, and La Mesa, respectively. We then used Tassel \cite{Bradbury et al 2007} to calculate linkage disequilibrium ($r^{2}$) across chromosome 1 for each population.

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \cite{Falush et al 2003} and the same phased 55K SNP data from \cite{Pyhajarvi et al 2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \cite{Cook et al 2012}. SNPs were anchored in a modified version of the IBM genetic map ( \cite{Gerke et al 2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped all SNPs within a 5 kb window into a haplotype. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

To investigate the phenotypic effects of the Hopscotch insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \cite{Hufford 2010} where the Hopscotch is segregating at highest frequency (0.4375) in our initial genotyping sample set. In order to maximize the likelihood of finding the Hopscotch in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer.

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \cite{Briggs et al 2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the Hopscotch insertion on plant architecture in maize \cite{Briggs et al 2007}. Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. Hopscotch genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the Hopscotch insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the Hopscotch on tillering index.

@@

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \cite{Fox and Weisberg 2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between Hopscotch genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS.

@@

For phenotyping 2 we measured tillering index every five days through day 50 for 302 plants. We followed the same transformations and analyses as described in phenotyping 1. We find the Hopscotch allele segregating at a frequency of 0.69, with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We find similar patterns as in phenotyping 1, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p=0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.

>>>>

\documentclass[12pt]{article}
\usepackage{geometry} 
\usepackage{hyperref}
\usepackage{color}
\usepackage[pdftex]{graphicx}  
%\usepackage{caption}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{subscript}
\renewcommand{\baselinestretch}{1.5}
\bibliographystyle{apalike}

\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} }
\newcommand{\mbh}[1]{\textcolor{red}{ \emph{\scriptsize  #1}} }
\newcommand{\lev}[1]{\textcolor{green}{\emph{\scriptsize #1}} }

@@

\author[1]{Laura Vann}
\author[1,2]{Thomas Kono}
\author[1,3]{Tanja Pyh\"aj\"arvi}
\author[1,4]{Matthew B. Hufford}
\author[1,5]{Jeffrey Ross-Ibarra\thanks{rossibarra@ucdavis.edu}}
\affil[1]{Department of Plant Sciences, University of California Davis}
\affil[2]{Department of Agronomy and Plant Genetics, University of Minnesota Twin Cities}
\affil[3]{Department of Biology, University of Oulu}
\affil[4]{Department of Ecology, Evolution, and Organismal Biology, Iowa State University}
\affil[5]{Center for Population Biology and Genome Center, University of California Davis}

@@

@@

The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Further studies have shown that the insertion of a transposable element (Hopscotch) upstream of \emph{tb1} enhances its expression, causing the reduction in branching observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of the age of insertion of the Hopscotch element, led us to investigate the role of \emph{tb1} in teosinte. Results from genotyping across many natural populations suggest that the Hopscotch element is segregating at a higher than expected frequency in a number of populations of both subspecies \emph{parviglumis} and subspecies \emph{mexicana}. Analysis of linkage disequilibrium between the Hopscotch element and variation (SNPs) in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte, and we find no evidence of environmental correlations that might suggest recent selection. Finally, two greenhouse experiments fail to find an important role for \emph{tb1} in controlling tillering in natural populations of \emph{parviglumis}. Our findings suggest that the role of the Hopscotch in tillering in teosinte is not as straightforward as it is in domesticated maize, and other loci known to affect branching architecture in maize may play a more important role in tillering in teosinte.

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. Perhaps the central focus of the study of domestication has been the identification of genetic variation underlying agronomically important traits such as fruit size and plant architecture \citep{OlsenGross2010}. Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \cite{pKovachMcCouch 2008}. But while some effort has been invested into understanding how wild alleles behave in their domesticated relatives, very little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors.

Maize (\emph{Zea mays} ssp. \emph{mays}) was domesticated from the teosinte \emph{Zea mays} ssp. \emph{parviglumis} (hereafter, \emph{parviglumis}) roughly 10,000 B.P. in southwest Mexico \citep{Piperno2009, Matsuoka2002}. Domesticated maize and the teosintes are an attractive system in which to study domestication due to the abundance of genetic tools developed for maize and well-characterized domestication loci \citep{Hufford2012a, Doebley2004, Hufford2012b}. Additionally, large naturally occurring populations of both \emph{Zea mays} ssp. \emph{parviglumis} (the wild progenitor of maize) and \emph{Zea mays} ssp. \emph{mexicana} (highland teosinte) can be found throughout Mexico \citep{Wilkes1977, Hufford2013}, and genetic diversity and effective population size (Ne) of these taxa is estimated to be high \citep{Ross-Ibarra2009} relative to other plants \citep{Nybom2004}.

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis for these changes has been a focus of maize research for a number of years. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. Further studies have shown that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. Tillering, or more specifically lack of tillers in teosinte, may provide an ecological advantage to plants. In species such as teosinte, where plants grow densely packed together, if plants invest more resources into growing taller and fewer resources into tillering, plants may be able to out-compete neighbors for limiting resources such as light. In natural populations of teosinte variation at the \emph(tb1) locus may then play a role in the ecology of teosinte.  In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{mexicana}, \emph{parviglumis}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

@@

We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{Zea mays} ssp. \emph{mexicana} (hereafter \emph{mexicana}) populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the Hopscotch insertion (\ref{TableS1Mays}, \ref{TableS2}). DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \cite{Kearse2012} to amplify the entire Hopscotch element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence (~5 kb amplification product) and absence (~300 bp amplification product) of the Hopscotch insertion. Two PCRs were performed for each individual, one with primers flanking the Hopscotch (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, 5'-TCGTTGATGCTTTGATGGATGG-3'; 
Hop R, 5'-AACAGTATGATTTCATGGGACCG-3'; and HopIntR, 5'-CCTCCACCTCTCATGAGATCC-3' (\ref{FigS1Locus}). Homozygotes show a single band for either the Hopscotch element (~5 kb) or the absence of the element (~300 bp), while heterozygotes are three-banded, showing a band for both the presence and absence of the Hopscotch element as well as a band for the internal primer set (\ref{FigS2Gel}). When only one PCR resolved well, we scored one allele for the individual, which explains the odd number of alleles included in our analyses. We used Phusion High Fidelity Enzyme (Finnzymes, Inc.) and the following conditions for amplifications: 98°C for 3 min, 30 cycles of 98°C for 15 s, 65°C for 30 s, and 72°C for 3 min 30 s, with a final extension of 72°C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the Hopscotch based on band size.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of the start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \citep{Wang1999, Clark2006}. We designed the following primers using PRIMER3 \citep{RozenSkaletsky2000}: for the 5' UTR, 5' GGATAATGTGCACCAGGTGT 3' and 5' GCGTGCTAGAGACACYTGTTGCT 3'; for the 50 kb upstream region, 5' TGTCCTCGCCGCAACTC 3' and 5' TGTACGCCCGCCCCTCATCA 3' (\ref{FigS1Locus}). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94°C for 3 min, 30 cycles of 92°C for 40 s, annealing for 1 min, 72°C for 40 s, and a final 10 min extension at 72°C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7°C and 58.8°C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 µL of amplification product. This mix was placed on a thermal cycler with the following program: 37°C for 30 min, 80°C for 15 min, and a final cool-down step to 4°C.

We cloned cleaned fragments into a TOPO-TA vector (Invitrogen, Carlsbad) using OneShot TOP10 chemically competent \emph{E. coli} cells, with an extended ligation time of 30 min for a complex target fragment. We plated cells on LB agar plates containing kanamycin, and screened colonies using vector primers M13 Forward and M13 Reverse under the following conditions: 96°C for 5 min; then 35 cycles at 96°C for 30 s, 53°C for 30 s, 72°C for two min; and a final extension at 72°C for 4 min. We visualized amplification products for incorporation of our insert on a 1\% agarose TAE gel.

Amplification products with successful incorporation of our insert were cleaned using Exonuclease 1 and Antarctic Phosphatase following the procedures detailed above, and sequenced with vector primers M13 Forward and M13 Reverse using Sanger sequencing at the College of Agriculture and Environmental Sciences (CAES) sequencing center at UC Davis. We aligned and trimmed primer sequences from resulting sequences using the software Geneious \citep{Kearse2012}. Following alignment, we verified singleton SNPs by sequencing an additional one to four colonies from each clone. If the singleton was not present in these additional sequences it was considered an amplification or cloning error, and we replaced the base with the base of the additional sequences. If the singleton appeared in at least one of the additional sequences we considered it a real variant and kept it for further analyses.

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE), and to calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \citep{Goudet2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \citep{Coop2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\citep{Coop2010}. We used genotyping and covariance data from Pyh\"aj\"arvi et al., (2013) \citep{Pyh\"aj\"arvi2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (\ref{TableS3BayEnv}). Environmental data were obtained from www.worldclim.org, the Harmonized World Soil Database and www.harvestchoice.org, and summarized by principle component analysis \citet{Pyh\"aj\"arvi2013} \citep{Pyh\"aj\"arvi2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \citep{Thornton2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\theta_{\subscript{W}}$; pairwise nucleotide diversity $\theta_{\subscript{\pi}}$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (http://evolution.genetics.washington.edu/phylip.html) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \citep{Chia2012}).

In order to assess patterns of linkage disequilibrium (LD) around the Hopscotch element in the context of chromosomal patterns of LD we used Tassel \citep{Bradbury2007} and calculated LD between SNPs across chromosome 1 using previously published data from twelve plants each of the Ejutla A, Ejutla B, San Lorenzo, and La Mesa populations \citep{Pyhajarvi2013}. We chose these populations because we had both genotyping data for the Hopscotch as well as chromosome-wide SNP data for chromosome 1. For each population we filtered the initial set of 5,897 SNPs on chromosome 1 to accept only SNPs with a minor allele frequency of at least 0.1, resulting in 1,671, 3,023, 3,122, and 2,167 SNPs for San Lorenzo, Ejutla B, Ejutla A, and La Mesa, respectively. We then used Tassel \citep{Bradbury2007} to calculate linkage disequilibrium ($r^{2}$) across chromosome 1 for each population.

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and the same phased 55K SNP data from \citep{Pyh\"aj\"arvi2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map (\citep{Gerke2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

To investigate the phenotypic effects of the Hopscotch insertion in teosinte, we conducted an initial phenotyping trial (Phenotyping 1). We germinated 250 seeds of \emph{parviglumis} collected in Jalisco state, Mexico (population San Lorenzo) \citep{Hufford2010} where the Hopscotch is segregating at highest frequency (0.4375) in our initial genotyping sample set. In order to maximize the likelihood of finding the Hopscotch in our association population we selected seeds from sites where genotyped individuals were homozygous or heterozygous for the insertion. We chose between 10-13 seeds from each of 23 sampling sites. We treated seeds with fungicide and germinated them in petri dishes with filter paper. Successful germinations (206 individuals) were then planted into one gallon size pots with potting soil and randomly spaced one foot apart on greenhouse benches. Plants were watered three times a day with an automatic drip containing 10-20-10 fertilizer.

Starting on day 15, we measured tillering index, the ratio of the sum of tiller lengths to the height of the plant \citep{Briggs2007}. Tillering index has been shown to be the most effective way to observe the phenotypic effects of the Hopscotch insertion on plant architecture in maize \citep{Briggs2007}. Following initial measurements, we phenotyped plants for tillering index every 5 days through day 40, and then on day 50 and day 60. On day 65 we measured culm diameter between the third and fourth nodes of each plant. Culm diameter is not believed to be correlated with tillering index, or variation at \emph{tb1} (e.g. Hopscotch genotype). Following phenotyping we extracted DNA from all plants using a modified SDS extraction protocol (http://www.ars.usda.gov). We genotyped individuals for the Hopscotch insertion following the protocols listed above. Based on these initial data, we conducted a post hoc power analysis using data from day 40 of phenotyping 1, indicating that a minimum of 71 individuals in each genotype class are needed to detect the observed effect of the Hopscotch on tillering index.

@@

Resulting tillering index data for each genotype class did not meet the criteria for a repeated measures ANOVA, so we transformed the data using a Box-Cox transformation ($\alpha=0$) implemented in the car package in R \citep{FoxWeisberg2011} to improve the normality and homogeneity of variance among genotype classes.  We analyzed relationships between genotype and tillering index and tiller number using a repeated measures ANOVA through a general linear model function implemented in SAS v.9.3 (SAS Institute Inc., Cary, NC, USA). Additionally, in order to compare any association between Hopscotch genotype and tillering and associations at other presumably unrelated traits, we performed an ANOVA between culm diameter and genotype using the same general linear model in SAS.

@@

For phenotyping 2 we measured tillering index every five days through day 50 for 302 plants. We followed the same transformations and analyses as described in phenotyping 1. We find the Hopscotch allele segregating at a frequency of 0.69, with a 0.6 frequency of Hopscotch homozygotes, and a 0.2 frequency of both heterozygotes and homozygotes for the teosinte allele. We find similar patterns as in phenotyping 1, with a weak positive correlation between tillering index and Hopscotch genotype at day 40 (p<0.0611), with no significant correlation on any day. Similarly, relationships between Hopscotch genotype and tiller number, and hopscotch genotype and culm diameter are not significant.
